HO unification from object language to meta language

### 2024-03-19 paper/main.tex
appendix in separate file

We represent a logic program \emph{run} in \Fo as
a list \emph{steps} $p$ of length $\mathcal{N}$. Each made of a
unification problem between terms $\mathcal{S}_{p_l}$ and
$\mathcal{S}_{p_r}$ taken from the set of all terms $\mathcal{S}$.
The composition of these steps starting from the
empty substitution $\rho_0$ produces the final
substitution $\rho_\mathcal{N}$.
\footnote{If the same rule is used multiple time in a run we
just consider as many copies as needed of the terms composing the
rules, with fresh unification variables each time}
The initial here $\rho_0$ is the empty substitution
%
\newcommand{\solve}{\ensuremath{\mathrm{check}}}
\newcommand{\fstep}{\ensuremath{\mathrm{fstep}}}
\newcommand{\hstep}{\ensuremath{\mathrm{hstep}}}
\newcommand{\frun}{\ensuremath{\mathrm{frun}}}
\newcommand{\hrun}{\ensuremath{\mathrm{hrun}}}
\newcommand{\stepF}[4]{\ensuremath{\fstep(#1,#2,#3) \mapsto #4}}
\newcommand{\stepFD}[5]{%
\ensuremath{#3 #1_{#2_l} \Uo #3 #1_{#2_r} \mapsto #4 \land #5 = #3 \cup #4}}
\newcommand{\stepH}[5]{\ensuremath{\hstep(#1,#2,#3,#4) \mapsto #5}}
\newcommand{\stepHD}[6]{\ensuremath{%
#3 #1_{#2_l} \Ue #3 #1_{#2_r} \mapsto #4 \land \solve(#6,#3 \cup #4) \mapsto #5}}
\newcommand{\runF}[3]{\ensuremath{\frun(#1,#2) \mapsto #3_{#2}}}
\newcommand{\runFD}[2]{\ensuremath{%
\bigwedge_{p = 1}^{#2} \stepF{#1}{p}{\rho_{p-1}}{\rho_{p}}}}
\newcommand{\runH}[3]{\ensuremath{\hrun(#1,#2) \mapsto #3_{#2}}}
\newcommand{\runHD}[3]{\ensuremath{%
\bigwedge_{p = 1}^{#2} \stepH{#1}{p}{\sigma_{p-1}}{#3}{\sigma_{p}}}}
\newcommand{\deff}{\ensuremath{\stackrel{{\scriptscriptstyle def}}{=\!=\!\!\!=}}}
$$
\begin{array}{l}
\stepF{\mathcal{S}}{p}{\rho}{\rho''}
\deff
\stepFD{\mathcal{S}}{p}{\rho}{\rho'}{\rho''}\vspace{2pt}\\
\runF{\mathcal{S}}{\mathcal{N}}{\rho}
\deff
\runFD{\mathcal{S}}{\mathcal{N}}
\end{array}
$$

@@

\subsection{The intuition in a nutshell}
A term $s$ is compiled in a term $t$ where any
``problematic'' subterm $tp$ is replaced by a unification variable $h$
and an accessory link, that represent a suspended unification problem
$h \Ue tp$. As a result \Ue is well behaved on $t$, that is it captures \Eo.
We now define ``problematic'' formally.

@@

\noindent
We call \elpiIn{fsubst} the memory of \Fo{}, while we call \elpiIn{subst}
the one of \Ho.
Both have the invariant that they are not cyclic, TODO explain.

@@

\otext{The following goal: \elpiIn{X = lam x\ x, app[X, 3] = 3}}
\otext{We use links-beta}

@@

\clearpage
\section*{Appendix}

Note that \elpiIn{(a infix b) c d} de-sugars to \elpiIn{(infix) a b c d}.

\begin{figure*}
\begin{elpicode}
~\PYG{k+kd}{type} \PYG{n+nf}{(\Eo)} \PYG{k+kt}{ftm -> ftm -> o}~.                           ~($\customlabel{pred:fequal}{\Eo}$)~
fapp A ~\Eo~fapp B :- map ~(\Eo)~ A B.
flam F ~\Eo~flam G :- pi x\ x ~\Eo~x => F x ~\Eo~G x.
fcon C ~\Eo~fcon C.
fuva N ~\Eo~fuva N.
flam F ~\Eo~T :-                                       ~($\customlabel{clause:eta1}{\eta_l}$)~
  pi x\ beta T [x] (R x), x ~\Eo~x => F x ~\Eo~R x.
T ~\Eo~flam F :-                                       ~($\customlabel{clause:eta2}{\eta_r}$)~
  pi x\ beta T [x] (R x), x ~\Eo~x => R x ~\Eo~F x.
fapp [flam X | L] ~\Eo~T :- beta (flam X) L R, R ~\Eo~T. ~($\customlabel{clause:beta1}{\beta_l}$)~
T ~\Eo~fapp [flam X | L] :- beta (flam X) L R, T ~\Eo~R. ~($\customlabel{clause:beta2}{\beta_r}$)~

type beta fm -> list fm -> fm -> o.
beta A [] A.
beta (flam F) [H | L] R :- subst F H B,
  beta B L R. % since F could be x\app[x|_] and H be lam _ 
beta (fapp A) L (fapp X) :- append A L X.
beta (fuva N) L (fapp [fuva N | L]).
beta (fcon H) L (fapp [fcon H | L]).

type subst (fm -> fm) -> fm -> fm -> o.
subst F H B :- napp (F H) B. % since (F H) may generate (app[app _|_])

type napp fm -> fm -> o.
napp (fcon C) (fcon C).
napp (flam F) (flam G) :- pi x\ napp x x => napp (F x) (G x).
napp (fapp[fapp L|M]) R :- !, append L M N, napp (fapp N) R.
napp (fapp[X]) R :- !, napp X R.
napp (fapp A) (fapp B) :- map napp A B.
napp (fuva N) (fuva N).
\end{elpicode}
\caption{Full implementation of the \Eo predicate for \Fo}
\label{code:eo-pred}
\Description[code:eo-pred]{code:eo-pred}
\end{figure*}

\begin{figure*}
\begin{elpicode}
~\PYG{k+kd}{type} \PYG{n+nf}{(\Ue)} \PYG{k+kt}{tm -> tm -> subst -> subst -> o}~.
% Congruence
(app A ~\Ue~app B) R S :- fold2 ~(\Ue)~A B R S.
(lam F ~\Ue~lam G) R S :- pi x\ (pi S\ (x ~\Ue~x) S S) => (F x ~\Ue~G x) R S.
~\Ue~(con C) (con C) S S.
% deref
(uva N A ~\Ue~T) R S :- set? N S F, move F A T1, (T1 ~\Ue~T) R S.
(T ~\Ue~uva N A) R S :- set? N S F, move F A T1, (T ~\Ue~T1) R S.
% flex-flex
(uva N A ~\Ue~uva M B) S S3 :-  unset? M, unset? N,
  distinct A, distinct B, 
  new S W S1, prune W Args1 B Ass, 
  assign N S1 Ass S2, assign M S2 Ass S3.
% assignment
(uva N A ~\Ue~T) R S :- distinct A, not (T = uva _ _), not_occ N S T,  
  bind A T T1, assign N S T1 S1.
(T ~\Ue~uva N A) R S :- distinct A, not (T = uva _ _), not_occ N S T, 
  bind A T T1, assign N S T1 S1.

type distinct list A -> o.
distinct [].
distinct [X|XS] :- name X, not(mem X XS),
distinct XS.

typeabbrev memory A (list (option A)).
type set? nat -> memory A -> A -> o.
set? N S T :- nth N S (some T).
type unset? nat -> memory A -> o.
unset? N S :- nth N S none.
type assign nat -> memory A -> A -> memory A -> o.
assign z [none|M] T [some T|M].
assign (s N) [X|M] T [X|M1] :- assign N M T M1. 
kind nat type.
type z nat.
type s nat -> nat.
type nth nat -> list A -> A -> o.
nth z [X|_] X.
nth (s N) [_|L] X :- nth N L X.

type new memory A -> nat -> memory A -> o.
new [] z [none].
new [X|XS] (s N) [X|YS] :- new XS N YS.

type prune .
type move .
type beta.
type bind.
type not_occ.
TODO

type fold2 (A -> A1 -> B -> B -> o) -> list A -> list A1 -> B -> B -> o.
fold2 _ [] [] A A.
fold2 F [X|XS] [Y|YS] A A1 :- F X Y A A0, fold2 F XS YS A0 A1.

\end{elpicode}
\caption{Implementation of the \Ue predicate for \Ho}
\label{code:ue-pred}
\Description[code:ue-pred]{code:ue-pred}
\end{figure*}

\begin{figure*}
\begin{elpicode}
type fold4 (A -> A1 -> B -> B -> C -> C -> o) -> list A -> list A1 -> B -> B -> C -> C -> o.
fold4 _ [] [] A A B B.
fold4 F [X|XS] [Y|YS] A A1 B B1 :- F X Y A A0 B B0, fold4 F XS YS A0 A1 B0 B1.

type len list A -> nat -> o.
len [] z.
len [_|L] (s X) :- len L X.

\end{elpicode}
\caption{Implementation of the compiler}
\label{code:ue-pred}
\Description[code:ue-pred]{code:ue-pred}
\end{figure*}

>>>>

We represent a logic program \emph{run} in \Fo as
a list \emph{steps} $p$ of length $\mathcal{N}$. Each made of a
unification problem between terms $\mathcal{S}_{p_l}$ and
$\mathcal{S}_{p_r}$ taken from the set of all terms $\mathcal{S}$.
The composition of these steps starting from the
empty substitution $\rho_0$ produces the final
substitution $\rho_\mathcal{N}$.
\footnote{If the same rule is used multiple time in a run we
just consider as many copies as needed of the terms composing the
rules, with fresh unification variables each time}
The initial here $\rho_0$ is the empty substitution
%
\newcommand{\solve}{\ensuremath{\mathrm{check}}\xspace}
\newcommand{\fstep}{\ensuremath{\mathrm{fstep}}\xspace}
\newcommand{\hstep}{\ensuremath{\mathrm{hstep}}\xspace}
\newcommand{\frun}{\ensuremath{\mathrm{frun}}\xspace}
\newcommand{\hrun}{\ensuremath{\mathrm{hrun}}\xspace}
\newcommand{\stepF}[4]{\ensuremath{\fstep(#1,#2,#3) \mapsto #4}}
\newcommand{\stepFD}[5]{%
\ensuremath{#3 #1_{#2_l} \Uo #3 #1_{#2_r} \mapsto #4 \land #5 = #3 \cup #4}}
\newcommand{\stepH}[5]{\ensuremath{\hstep(#1,#2,#3,#4) \mapsto #5}}
\newcommand{\stepHD}[6]{\ensuremath{%
#3 #1_{#2_l} \Ue #3 #1_{#2_r} \mapsto #4 \land \solve(#6,#3 \cup #4) \mapsto #5}}
\newcommand{\runF}[3]{\ensuremath{\frun(#1,#2) \mapsto #3_{#2}}}
\newcommand{\runFD}[2]{\ensuremath{%
\bigwedge_{p = 1}^{#2} \stepF{#1}{p}{\rho_{p-1}}{\rho_{p}}}}
\newcommand{\runH}[3]{\ensuremath{\hrun(#1,#2) \mapsto #3_{#2}}}
\newcommand{\runHD}[3]{\ensuremath{%
\bigwedge_{p = 1}^{#2} \stepH{#1}{p}{\sigma_{p-1}}{#3}{\sigma_{p}}}}
\newcommand{\deff}{\ensuremath{\stackrel{{\scriptscriptstyle def}}{=\!=\!\!\!=}}}
$$
\begin{array}{l}
\stepF{\mathcal{S}}{p}{\rho}{\rho''}
\deff
\stepFD{\mathcal{S}}{p}{\rho}{\rho'}{\rho''}\vspace{2pt}\\
\runF{\mathcal{S}}{\mathcal{N}}{\rho}
\deff
\runFD{\mathcal{S}}{\mathcal{N}}
\end{array}
$$

@@

\subsection{The intuition in a nutshell}
\label{sec:nutshell}
A term $s$ is compiled in a term $t$ where any
``problematic'' subterm $tp$ is replaced by a unification variable $h$
and an accessory link, that represent a suspended unification problem
$h \Ue tp$. As a result \Ue is well behaved on $t$, that is it captures \Eo.
We now define ``problematic'' formally.

@@

\noindent
We call \elpiIn{fsubst} the memory of \Fo{}, while we call \elpiIn{subst}
the one of \Ho.
Both have the invariant that they are not cyclic, TODO explain.
Other invariant: the terms in ho\_subst never contains eta and beta expansion

@@

$\beta$-reduction problems (\maybebeta) appears any time we deal with a subterm $t
= X t_1 \dots t_n$, where $X$ is flexible and the list $[t_1 \dots t_n]$ in not
in \llambda. This unification problem is not solvable without loss of
generality, since there is not a most general unifier. If we take back the
example given in \cref{sec:nutshell}, the unification $F a = a$ admits two solutions for $F$:
$\rho_1 = \{F \mapsto \lambda x.x\}$ and $\rho_2 = \{F \mapsto \lambda \_.a\}$.
Despite this, it is possible to work with $\maybebeta$ if an oracle provides a
substitution $\rho$ such that $\rho t$ falls again in the \llambda.

On the other hand, the \Ue is not designed to understand how the $\beta$-redexes
work in the onject language. Therefore, even if we know that $F$ is assigned
to $\lambda x.x$, \Ue is not able to unify $F a$ with $a$. On the other hand,
the problem $F a = G$ is solvable by \Ue, but the final result is that $G$ is
assigned to $(\lambda x.x) a$ which breaks the invariant saying that the 
substitution of the meta language always contain only terms in normal form.

\otext{The following goal: \elpiIn{X = lam x\ x, app[X, a] = a}}
\otext{We use links-beta}

@@

\clearpage

\input{appendix.tex}
